---
title: "Surveillance and outbreak response"
subtitle: "Where to sample, and what to do when something is already spreading"
description: "Biodiversity-informed wildlife disease surveillance, and earlier outbreak response work on COVID-19 and mpox."
categories: [Surveillance, One Health, Outbreak Response]
toc: true
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note appearance="minimal"}
Placeholder page. Framing text is a draft. This page absorbs the former
COVID-19 and mpox theme pages, which is why those sections are written in the
past tense. Explainer notes for the papers are in `data/paper_notes/`.
:::

Surveillance is a resource allocation problem before it is a laboratory one.
Sampling wildlife is expensive, so the question is where a sample changes what
we know. Outbreak response is the same question under time pressure.

## Wildlife disease surveillance

**A protocol for biodiversity-informed wildlife disease surveillance.**
*PLOS Ecosystems*, 2026. Host species distribution models set a biodiversity
dose, uncertainty identifies where discovery sampling would help most, and
balanced-acceptance sampling turns the result into field sites. Case studies on
*Hantaviridae* and *Arenaviridae* in India and South Korea.

Ongoing work extends this to outbreak settings, where the sampling frame has to
be revised as an event unfolds.

## mpox

Work from the 2022 outbreak and from the endemic range that preceded it.

**Increased outbreaks of mpox highlight gaps in actual disease burden.**
*International Journal of Infectious Diseases*, 2022. Two hypotheses for the
global outbreak: growing synanthropic rodent populations in West and Central
Africa, and an immunologically naive population 45 years after smallpox
vaccination ended.

**Community transmission of mpox in the United Kingdom.** *Eurosurveillance*,
2022. The first UK cases without travel history, described during the
investigation, from work in the UK Health Security Agency outbreak
surveillance team.

The reservoir of mpox virus is still unidentified. Virus has been isolated from
wild animals twice, and antibodies have been found in several further species.

```{r}
#| label: mpox-setup
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, kableExtra, here, readr)
```

``` {r monkeypox-table, echo = FALSE, message = FALSE, warning = FALSE}
monkeypox_table <- read_csv(here("monkeypox", "data", "monkeypox_reservoirs.csv"))

monkeypox_table %>%
  mutate(References = cell_spec(References, "html", link = Link)) %>%
  select(-Link, -Identification) %>%
  kable("html", escape = FALSE, caption = "Adapted from Haider et al. 2022") %>%
  pack_rows(index = table(fct_inorder(monkeypox_table$Identification))) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),
                font_size = 10)

```

## COVID-19

Between 2020 and 2022 I worked on the epidemiology of SARS-CoV-2: modelling
international spread with the LSHTM Centre for Mathematical Modelling of
Infectious Diseases, a living review of smoking and COVID-19 outcomes,
clinical work through the pandemic, and real-time surveillance of the Omicron
variant with the UK Health Security Agency.

That work is finished. The papers are listed on the
[publications page](/publications.html).

::: {.callout-note appearance="minimal"}
To add: the COVID-19 figure carried over from the old theme page, and a short
account of the UKHSA surveillance work.
:::
