---
title: "Lassa epidemiology"
subtitle: "Case burden, reporting and the limits of the sampling record"
description: "How many Lassa fever cases there really are, and what the surveillance and sequence record can support."
categories: [Lassa Fever, Epidemiology]
toc: true
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note appearance="minimal"}
Placeholder page. The framing text is a draft. The explainer notes for the
under-reporting and sequencing-bias papers are written and will be pulled in
from `data/paper_notes/` when the programme pages are wired to the
publications data.
:::

Reported Lassa fever case counts are the input to almost every risk model, and
they are incomplete in ways that vary by country and by year. This programme
asks what can be inferred anyway.

## Detections in hosts

Small mammal species in which *Mammarenavirus lassaense* has been detected in West Africa, by assay, with the studies that report each detection. The records come from the Project ArHa database, reviewed and extended to 2026.

```{r}
#| label: lassa-reservoirs

# --- 1. Load Packages ---
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, kableExtra, here, ggspatial, cowplot, readr)

# --- 2. Host detections ---
# data/lassa_hosts.csv is built by scripts/build_lassa_hosts.R. One row per
# species, assay, country and study reporting a positive animal.
lassa_hosts <- read_csv(here("data", "lassa_hosts.csv"), show_col_types = FALSE)

# One row per species and assay; the studies listed in order of publication,
# each linked once. Genera ordered by their earliest detection.
assay_order <- c("Virus Culture", "PCR", "Serology")

host_table <- lassa_hosts |>
  distinct(host_genus, host_species, assay, publication_year, reference, link) |>
  arrange(publication_year) |>
  summarise(first_year = min(publication_year),
            references = paste(cell_spec(reference, "html", link = link), collapse = "; "),
            .by = c(host_genus, host_species, assay)) |>
  mutate(genus_first = min(first_year), .by = host_genus) |>
  mutate(assay = factor(assay, levels = assay_order)) |>
  arrange(genus_first, host_genus, host_species, assay)

genus_rows <- table(fct_inorder(host_table$host_genus))

host_table |>
  mutate(host_species = paste0("<em>", host_species, "</em>"),
         assay = str_replace(as.character(assay), "Virus Culture", "Virus culture")) |>
  select(Species = host_species, Assay = assay, References = references) |>
  kbl("html", escape = FALSE,
      caption = "Small mammal species with reported <em>Mammarenavirus lassaense</em> detections in West Africa, by assay.") |>
  pack_rows(index = genus_rows) |>
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 10)
```

## Reported human cases

```{r lassa-map, echo = FALSE, message = FALSE, warning = FALSE}
lassa_raw <- read_csv(here("lassa", "data", "confirmed_cases.csv"))

lassa_clean <- lassa_raw %>%
  filter(case_when(country == "nigeria" ~ region == "all",
                   TRUE ~ confirmed_cases > 0)) %>%
  group_by(year, country) %>%
  summarise(confirmed_cases = sum(confirmed_cases)) %>%
  mutate(country = str_to_title(country),
         country = recode(country, 
                          "Cote D'ivoire" = "Côte d'Ivoire"),
         # confirmed_cases = log(confirmed_cases),
         endemic = factor(case_when(country %in% c("Benin", "Ghana", "Guinea", "Liberia", "Mali",
                                                   "Nigeria", "Sierra Leone", "Togo") ~ "",
                             TRUE ~ "Non-endemic")),
         country_label = factor(case_when(country %in% c("Benin", "Ghana", "Guinea", "Liberia", "Mali",
                                                         "Nigeria", "Sierra Leone", "Togo") ~ country,
                                          TRUE ~ "Combined"),
                                levels = c("Benin", "Ghana", "Guinea", "Liberia", "Mali",
                                           "Nigeria", "Sierra Leone", "Togo", "Combined"))) %>%
  filter(year >= 2011)

colours_wa <- c("Endemic countries" = "white",
                "Benin" = "#871a10",
                "Ghana" = "#000000",
                "Guinea" = "#be2a2c",
                "Liberia" = "#0a2864",
                "Mali" = "#50ab4a",
                "Nigeria" = "#398053",
                "Sierra Leone" = "#2f71c1",
                "Togo" = "#f7d147",
                " " = "white",
                "Non-endemic countries" = "white",
                "Côte d'Ivoire" = "#e6853b",
                "Burkina Faso" = "#107d87")

lassa_clean$country = factor(lassa_clean$country, levels=c("Endemic countries", "Benin", "Ghana", "Guinea",
                                                           "Liberia", "Mali", "Nigeria", "Sierra Leone", "Togo", " ",
                                                           "Non-endemic countries", "Côte d'Ivoire", "Burkina Faso"))

case_plot <- ggplot(lassa_clean, aes(x = year, y = confirmed_cases, fill = country)) +
  geom_col() +
  scale_x_continuous(breaks = seq(from = min(lassa_clean$year), to = max(lassa_clean$year), by = 2)) +
  scale_y_continuous(breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1))))) +
  scale_fill_manual(values = colours_wa,
                    drop=FALSE) +
  guides(fill=guide_legend(ncol=1)) +
  theme(legend.key = element_rect(fill=NA),
        legend.title = element_blank()) +
  facet_wrap(~ endemic + country_label , scales = "free_y") +
  annotate(geom = "rect", xmin = 2013, xmax = 2016, ymin = -Inf, ymax = Inf, alpha = 0.1) +
  annotate(geom = "rect", xmin = 2020, xmax = 2024, ymin = -Inf, ymax = Inf, alpha = 0.1) +
  annotate(geom = "rect", xmin = year(Sys.Date()), xmax = year(Sys.Date()) + 1, ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "orange") +
  theme_minimal() +
  labs(x = element_blank(),
       y = "Confirmed cases",
       fill = element_blank())

mastomys_data <- read_rds(here("lassa", "data", "mastomys_data.rds"))

mastomys_distribution <- ggplot() +
  geom_sf(data = mastomys_data$wa_countries, aes(fill = endemic), lwd = 0.1) +
  geom_sf_text(data = mastomys_data$wa_countries, aes(label = NAME_0), size = 5) +
  scale_fill_manual(values = c("Endemic" = "#f2aa86",
                               "Outbreaks" = "#fef8e1",
                               "No cases" = "#fafafa")) +
  geom_sf(data = mastomys_data$iucn, fill = "#C12D20", alpha = 0.4, lwd = 0.1) +
  theme_minimal() +
  labs(fill = element_blank(),
       x = element_blank(),
       y = element_blank()) +
  coord_sf(xlim = c(mastomys_data$afr_bbox[1], mastomys_data$afr_bbox[3]), ylim = c(mastomys_data$afr_bbox[2], mastomys_data$afr_bbox[4])) +
  annotation_scale(height = unit(0.1, "cm")) +
  annotation_north_arrow(style = north_arrow_minimal(text_size = 7),
                         location = "tl")

# This map has been produced using IUCN range data and country shapefiles. The code is available here: https://github.com/DidDrog11/miscellaneous/blob/main/scripts/lassa_cases.R

combined_plot <- plot_grid(case_plot, mastomys_distribution, nrow = 2,
                           labels = c("A", "B"), greedy = TRUE)
save_plot(here("lassa", "images", "lassa_cases_endemic.svg"), combined_plot, base_height = c(8, 6), base_width = c(12, 12))

```

![A) Reported human cases of Lassa fever, 2011 onwards. The first grey band (2013-2016) is the West African Ebola epidemic, the second (2020-) the COVID-19 pandemic, and the orange band an incomplete year of reporting. B) Predicted distribution of *Mastomys natalensis* over known endemic countries.](/lassa/images/lassa_cases_endemic.svg)

::: {.callout-warning appearance="minimal"}
The case figure is current to 2023-07-13 and needs regenerating.
:::

## Papers

```{r}
#| results: asis
#| echo: false
#| message: false
#| warning: false
# Papers assigned to this page in data/publications_overrides.csv.
source(here::here("scripts", "paper_cards.R"))
render_programme_papers("research/lassa-epidemiology.qmd")
```
