---
title: "Lassa epidemiology"
subtitle: "Case burden, reporting and the limits of the sampling record"
description: "How many Lassa fever cases there really are, and what the surveillance and sequence record can support."
categories: [Lassa Fever, Epidemiology]
toc: true
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note appearance="minimal"}
Placeholder page. The framing text is a draft. The explainer notes for the
under-reporting and sequencing-bias papers are written and will be pulled in
from `data/paper_notes/` when the programme pages are wired to the
publications data.
:::

Reported Lassa fever case counts are the input to almost every risk model, and
they are incomplete in ways that vary by country and by year. This programme
asks what can be inferred anyway.

## Detections in hosts

Small mammal species in which *Mammarenavirus lassaense* has been detected in West Africa, by assay, with the studies that report each detection. The records come from the Project ArHa database, reviewed and extended to 2026.

```{r}
#| label: lassa-reservoirs

# --- 1. Load Packages ---
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, kableExtra, here, ggspatial, cowplot, readr)

# --- 2. Host detections ---
# data/lassa_hosts.csv is built by scripts/build_lassa_hosts.R. One row per
# species, assay, country and study reporting a positive animal.
lassa_hosts <- read_csv(here("data", "lassa_hosts.csv"), show_col_types = FALSE)

# One row per species and assay; the studies listed in order of publication,
# each linked once. Genera ordered by their earliest detection.
assay_order <- c("Virus Culture", "PCR", "Serology")

host_table <- lassa_hosts |>
  distinct(host_genus, host_species, assay, publication_year, reference, link) |>
  arrange(publication_year) |>
  summarise(first_year = min(publication_year),
            references = paste(cell_spec(reference, "html", link = link), collapse = "; "),
            .by = c(host_genus, host_species, assay)) |>
  mutate(genus_first = min(first_year), .by = host_genus) |>
  mutate(assay = factor(assay, levels = assay_order)) |>
  arrange(genus_first, host_genus, host_species, assay)

genus_rows <- table(fct_inorder(host_table$host_genus))

host_table |>
  mutate(host_species = paste0("<em>", host_species, "</em>"),
         assay = str_replace(as.character(assay), "Virus Culture", "Virus culture")) |>
  select(Species = host_species, Assay = assay, References = references) |>
  kbl("html", escape = FALSE,
      caption = "Small mammal species with reported <em>Mammarenavirus lassaense</em> detections in West Africa, by assay.") |>
  pack_rows(index = genus_rows) |>
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 10)
```

## Reported human cases

Confirmed Lassa fever cases by country and year. Nigeria publishes weekly situation reports, so its series is complete from 2016 and current to the latest report. Figures for the other countries are compiled by hand from WHO, ECDC, ProMED, national reports and publications, and are often partial. Each panel gives the latest year with data. Newer figures or corrections are welcome.

```{r}
#| label: lassa-cases
#| fig-width: 10
#| fig-height: 7.5
#| dev.args: !expr list(bg = "transparent")

# data/lassa_cases.csv is built by scripts/build_lassa_cases.R.
lassa_cases <- read_csv(here("data", "lassa_cases.csv"), show_col_types = FALSE)

# Panel label carries the data vintage, so a stale country cannot pass as
# current. Nigeria first, then countries by total cases.
panel_labels <- lassa_cases |>
  slice_max(year, by = country) |>
  mutate(label = if_else(coverage == "" | is.na(coverage),
                         paste0(country, ", to ", year),
                         paste0(country, ", ", coverage, " ", year))) |>
  select(country, label)

panel_order <- lassa_cases |>
  summarise(total = sum(confirmed_cases), .by = country) |>
  arrange(country != "Nigeria", desc(total)) |>
  left_join(panel_labels, by = "country")

case_data <- lassa_cases |>
  left_join(panel_labels, by = "country") |>
  mutate(label = factor(label, levels = panel_order$label),
         year_status = factor(if_else(complete, "Complete year", "Partial year"),
                              levels = c("Complete year", "Partial year")))

# Context bands: the West African Ebola epidemic and the COVID-19 pandemic,
# both of which disrupted Lassa fever surveillance.
context_bands <- tibble(xmin = c(2013.5, 2019.5), xmax = c(2016.5, 2022.5))

ggplot(case_data, aes(x = year, y = confirmed_cases, fill = year_status)) +
  geom_rect(data = context_bands, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "grey50", alpha = 0.12) +
  geom_col(width = 0.8) +
  facet_wrap(~ label, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Complete year" = "#13563b", "Partial year" = "#53a57f")) +
  scale_x_continuous(limits = c(2007.5, 2026.5), breaks = seq(2008, 2026, by = 4)) +
  scale_y_continuous(breaks = function(x) unique(floor(pretty(c(0, x[2]), n = 3)))) +
  labs(x = NULL, y = "Confirmed cases", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        strip.text = element_text(hjust = 0, face = "bold"),
        plot.background = element_blank(), panel.background = element_blank())
```

Grey bands mark the West African Ebola epidemic (2014 to 2016) and the COVID-19 pandemic (2020 to 2022), both of which disrupted surveillance. [Download the data (CSV)](/data/lassa_cases.csv), with the source of every value.

```{r}
#| label: lassa-case-sources
#| results: asis

# One row per country: coverage and the kinds of source behind it.
case_sources <- lassa_cases |>
  mutate(source = str_remove(source, " \\(.*\\)$")) |>
  summarise(Years = paste(min(year), "to", max(year)),
            `Country-years` = n(),
            Sources = paste(sort(unique(unlist(str_split(source, "; ")))), collapse = ", "),
            .by = country) |>
  left_join(panel_order |> select(country, total), by = "country") |>
  arrange(country != "Nigeria", desc(total)) |>
  select(Country = country, Years, `Country-years`, Sources)

cat('<details class="globe-table"><summary>Show sources by country</summary>\n\n')
case_sources |>
  kbl("html", caption = "Years covered and sources for each country.") |>
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 12) |>
  cat()
cat("\n</details>\n")
```

```{r}
#| label: mastomys-range
#| fig-width: 10
#| fig-height: 6
#| dev.args: !expr list(bg = "transparent")
mastomys_data <- read_rds(here("lassa", "data", "mastomys_data.rds"))

ggplot() +
  geom_sf(data = mastomys_data$wa_countries, aes(fill = endemic), lwd = 0.1) +
  geom_sf_text(data = mastomys_data$wa_countries, aes(label = NAME_0), size = 4) +
  scale_fill_manual(values = c("Endemic" = "#f2aa86", "Outbreaks" = "#fef8e1", "No cases" = "#fafafa")) +
  geom_sf(data = mastomys_data$iucn, fill = "#C12D20", alpha = 0.4, lwd = 0.1) +
  coord_sf(xlim = c(mastomys_data$afr_bbox[1], mastomys_data$afr_bbox[3]),
           ylim = c(mastomys_data$afr_bbox[2], mastomys_data$afr_bbox[4])) +
  annotation_scale(height = unit(0.1, "cm")) +
  annotation_north_arrow(style = north_arrow_minimal(text_size = 7), location = "tl") +
  labs(fill = NULL, x = NULL, y = NULL) +
  theme_minimal() +
  theme(plot.background = element_blank(), panel.background = element_blank())
```

The IUCN range of *Mastomys natalensis* (red) over countries with endemic Lassa fever, countries with outbreaks, and countries with no reported cases.

## Papers

```{r}
#| results: asis
#| echo: false
#| message: false
#| warning: false
# Papers assigned to this page in data/publications_overrides.csv.
source(here::here("scripts", "paper_cards.R"))
render_programme_papers("research/lassa-epidemiology.qmd")
```
