---
title: "Lassa epidemiology"
subtitle: "Case burden, reporting and the limits of the sampling record"
description: "How many Lassa fever cases there really are, and what the surveillance and sequence record can support."
categories: [Lassa Fever, Epidemiology]
toc: true
execute:
  echo: false
  warning: false
  message: false
---

Reported Lassa fever case counts feed almost every risk model, yet they are incomplete in ways that vary by country and by year. This programme assesses what can be inferred from the surveillance, sampling and sequence record as it stands: which hosts carry the virus, how far human cases are under-reported, and what a change in a reported indicator can and cannot show.


## Detections in hosts

Small mammal species in which *Mammarenavirus lassaense* has been detected in West Africa, by assay, with the studies that report each detection. The records come from the Project ArHa database, reviewed and extended to 2026.

```{r}
#| label: lassa-reservoirs

# --- 1. Load Packages ---
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, kableExtra, here, sf, readr)

# --- 2. Host detections ---
# data/lassa_hosts.csv is built by scripts/build_lassa_hosts.R. One row per
# species, assay, country and study reporting a positive animal.
lassa_hosts <- read_csv(here("data", "lassa_hosts.csv"), show_col_types = FALSE)

# One row per species and assay; the studies listed in order of publication,
# each linked once. Genera ordered by their earliest detection.
assay_order <- c("Virus Culture", "PCR", "Serology")

host_table <- lassa_hosts |>
  distinct(host_genus, host_species, assay, publication_year, reference, link) |>
  arrange(publication_year) |>
  summarise(first_year = min(publication_year),
            references = paste(cell_spec(reference, "html", link = link), collapse = "; "),
            .by = c(host_genus, host_species, assay)) |>
  mutate(genus_first = min(first_year), .by = host_genus) |>
  mutate(assay = factor(assay, levels = assay_order)) |>
  arrange(genus_first, host_genus, host_species, assay)

genus_rows <- table(fct_inorder(host_table$host_genus))

host_table |>
  mutate(host_species = paste0("<em>", host_species, "</em>"),
         assay = str_replace(as.character(assay), "Virus Culture", "Virus culture")) |>
  select(Species = host_species, Assay = assay, References = references) |>
  kbl("html", escape = FALSE,
      caption = "Small mammal species with reported <em>Mammarenavirus lassaense</em> detections in West Africa, by assay.") |>
  pack_rows(index = genus_rows) |>
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 10)
```

## Reported human cases

Confirmed Lassa fever cases by country and year. Nigeria publishes weekly situation reports, so its series is complete from 2016 and current to the latest report. Figures for the other countries are compiled by hand from WHO, ECDC, ProMED, national reports and publications, and are often partial. Each panel gives the latest year with data. Newer figures or corrections are welcome.

```{r}
#| label: lassa-cases
#| fig-width: 10
#| fig-height: 7.5
#| dev.args: !expr list(bg = "transparent")

# data/lassa_cases.csv is built by scripts/build_lassa_cases.R.
lassa_cases <- read_csv(here("data", "lassa_cases.csv"), show_col_types = FALSE)

# Panel label carries the data vintage, so a stale country cannot pass as
# current. Nigeria first, then countries by total cases.
panel_labels <- lassa_cases |>
  slice_max(year, by = country) |>
  mutate(label = if_else(coverage == "" | is.na(coverage),
                         paste0(country, ", to ", year),
                         paste0(country, ", ", coverage, " ", year))) |>
  select(country, label)

panel_order <- lassa_cases |>
  summarise(total = sum(confirmed_cases), .by = country) |>
  arrange(country != "Nigeria", desc(total)) |>
  left_join(panel_labels, by = "country")

case_data <- lassa_cases |>
  left_join(panel_labels, by = "country") |>
  mutate(label = factor(label, levels = panel_order$label),
         year_status = factor(if_else(complete, "Complete year", "Partial year"),
                              levels = c("Complete year", "Partial year")))

# Context bands: the West African Ebola epidemic and the COVID-19 pandemic,
# both of which disrupted Lassa fever surveillance.
context_bands <- tibble(xmin = c(2013.5, 2019.5), xmax = c(2016.5, 2022.5))

ggplot(case_data, aes(x = year, y = confirmed_cases, fill = year_status)) +
  geom_rect(data = context_bands, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "grey50", alpha = 0.12) +
  geom_col(width = 0.8) +
  facet_wrap(~ label, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Complete year" = "#13563b", "Partial year" = "#53a57f")) +
  scale_x_continuous(limits = c(2007.5, 2026.5), breaks = seq(2008, 2026, by = 4)) +
  scale_y_continuous(breaks = function(x) unique(floor(pretty(c(0, x[2]), n = 3)))) +
  labs(x = NULL, y = "Confirmed cases", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        strip.text = element_text(hjust = 0, face = "bold"),
        plot.background = element_blank(), panel.background = element_blank())
```

Grey bands mark the West African Ebola epidemic (2014 to 2016) and the COVID-19 pandemic (2020 to 2022), both of which disrupted surveillance. [Download the data (CSV)](/data/lassa_cases.csv), with the source of every value.

```{r}
#| label: lassa-case-sources
#| results: asis

# One row per country: coverage and the kinds of source behind it.
case_sources <- lassa_cases |>
  mutate(source = str_remove(source, " \\(.*\\)$")) |>
  summarise(Years = paste(min(year), "to", max(year)),
            `Country-years` = n(),
            Sources = paste(sort(unique(unlist(str_split(source, "; ")))), collapse = ", "),
            .by = country) |>
  left_join(panel_order |> select(country, total), by = "country") |>
  arrange(country != "Nigeria", desc(total)) |>
  select(Country = country, Years, `Country-years`, Sources)

cat('<details class="globe-table"><summary>Show sources by country</summary>\n\n')
case_sources |>
  kbl("html", caption = "Years covered and sources for each country.") |>
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 12) |>
  cat()
cat("\n</details>\n")
```

```{r}
#| label: mastomys-range
#| fig-width: 10
#| fig-height: 6
#| dev.args: !expr list(bg = "transparent")
mastomys_data <- read_rds(here("lassa", "data", "mastomys_data.rds"))

# Country status is ordinal, so one hue from light to dark (checked with the
# dataviz palette validator against the page ground). The host range is drawn
# as an outline, not a fill, so it never mixes with the country colours.
status_levels <- c("Endemic", "Outbreaks", "No cases")
status_fill <- c("Endemic" = "#3a654f", "Outbreaks" = "#6b8f78", "No cases" = "#9db2a3")

countries <- mastomys_data$wa_countries |>
  mutate(endemic = factor(endemic, levels = status_levels),
         label_colour = if_else(endemic == "Endemic", "#ffffff", "#1e2420"))

# Small coastal countries are left unlabelled; their names would overlap.
# Sierra Leone is nudged inland so its white label does not run onto the sea.
country_labels <- countries |>
  filter(!NAME_0 %in% c("Gambia", "Guinea-Bissau")) |>
  st_point_on_surface()
label_xy <- st_coordinates(country_labels)
country_labels <- country_labels |>
  st_drop_geometry() |>
  mutate(x = label_xy[, 1] + if_else(NAME_0 == "Sierra Leone", 0.6, 0), y = label_xy[, 2])

# The boundary only, clipped by the map frame, so no crop edges are drawn.
host_range <- mastomys_data$iucn |>
  st_make_valid() |>
  st_boundary() |>
  mutate(layer = "IUCN range of Mastomys natalensis")

ggplot() +
  geom_sf(data = countries, aes(fill = endemic), colour = "#f2f4ef", linewidth = 0.4) +
  geom_sf(data = host_range, aes(colour = layer), linewidth = 1) +
  geom_text(data = country_labels, aes(x = x, y = y, label = NAME_0, colour = I(label_colour)), size = 3.6) +
  scale_fill_manual(values = status_fill, name = "Lassa fever") +
  scale_colour_manual(values = c("IUCN range of Mastomys natalensis" = "#7e6458"), name = NULL,
                      labels = expression(paste("IUCN range of ", italic("Mastomys natalensis")))) +
  coord_sf(xlim = c(mastomys_data$afr_bbox[1], mastomys_data$afr_bbox[3]),
           ylim = c(mastomys_data$afr_bbox[2], mastomys_data$afr_bbox[4]), expand = FALSE) +
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2)) +
  theme_void(base_size = 12) +
  theme(legend.position = "right", legend.justification = "top",
        plot.background = element_blank(), panel.background = element_blank())
```

Countries by Lassa fever status, with the IUCN range of *Mastomys natalensis*, the main reservoir, outlined.

## Estimated burden

Reported cases are a small fraction of infections. The [socio-economic shield paper](/research/lassa-cities.html) estimates the regional burden from where the reservoir persists, including at the urban fringe, and how urban infrastructure dampens spillover. After adjusting for seroreversion and shielding, it estimates about 2.6 million Lassa virus infections a year across the endemic region. It also identifies districts in Nigeria, Benin and Togo with high predicted infection and no reported cases.

![**Silent districts.** A, B) Districts in Nigeria and the Mano River region by confirmed endemic status and predicted risk where no cases are reported. C) Predicted annual infections against reported cases by district. From the [socio-economic shield paper](/research/lassa-cities.html).](../lassa/images/silent_districts_map.png){fig-alt="Map of West African districts, highlighting those with high predicted Lassa virus infections and no reported cases, in Nigeria, Benin and Togo." width="80%"}

## Ongoing work

::: {.pub-entry .pub-entry-plain #work-lassa-cfr}
::: {.pub-body}
**Reported Lassa fever case fatality in Nigeria, 2017–2026**  
[Analysis plan, 2026](https://doi.org/10.5281/zenodo.23135368) · [Code on GitHub](https://github.com/DidDrog11/lassa-cfr)

<details class="pub-more">
<summary>Summary</summary>

::: {.pub-more-body}
The Nigeria Centre for Disease Control and Prevention reports a case fatality rate among laboratory-confirmed Lassa fever cases in every weekly situation report. A reported case fatality rate is a ratio of two surveillance counts, so it can change through who is tested, which deaths are found, where cases occur and when the figure is read, as well as through a change in the risk of death.

The study sets out these mechanisms, assesses which of them the published situation reports from 2017 to 2026 allow to be evaluated, and evaluates those that can be. The analysis plan was fixed and archived before the analyses were run on the final data. No results are available yet.
:::

</details>
:::
:::

::: {.pub-entry .pub-entry-plain #work-lassa-chapter}
::: {.pub-body}
**Lassa fever**  
Simons D, Bangura U, Fichet-Calvet E, Asogun D, Friant S  
Chapter 1.29 in Kock R (ed.), *Zoonoses and Zoonotic Diseases: A Comprehensive Reference*, Elsevier. Forthcoming.

A One Health review of the virus, its reservoir, epidemiology, clinical disease, diagnosis and treatment, and the integrated strategies needed for its control.
:::
:::

## Data and code

- **Under-reporting.** Code and data for the estimate of under-reported cases from reported deaths are on [GitHub](https://github.com/DidDrog11/lassa_underreporting).
- **Hosts and cases.** The [host detections](/data/lassa_hosts.csv) and [confirmed cases](/data/lassa_cases.csv) behind the table and figure on this page.

## Papers

```{r}
#| results: asis
#| echo: false
#| message: false
#| warning: false
# Papers assigned to this page in data/publications_overrides.csv.
source(here::here("scripts", "paper_cards.R"))
render_programme_papers("research/lassa-epidemiology.qmd")
```
