---
title: "Surveillance and outbreak response"
subtitle: "Where to sample, and what to do when something is already spreading"
categories: [Surveillance, One Health, Outbreak Response]
toc: true
execute:
  echo: false
  warning: false
  message: false
---

Surveillance is a resource allocation problem before it is a laboratory one. Sampling wildlife is expensive, so the question is where a sample changes what we know. Outbreak response is the same question under time pressure.

```{r}
#| label: setup
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, kableExtra, here)
source(here("scripts", "paper_cards.R"))
this_page <- "research/surveillance.qmd"
```

## Wildlife disease surveillance

A [protocol for biodiversity-informed wildlife disease surveillance](https://doi.org/10.1371/journal.pesy.0000023) sets out where to sample. Host species distribution models set a biodiversity dose, uncertainty identifies where discovery sampling would help most, and balanced-acceptance sampling turns the result into field sites. The case studies are *Hantaviridae* and *Arenaviridae* in India and South Korea.

The Vetenskapsrådet South Korea–Sweden research collaboration grant *An AI-directed One Health risk prediction framework for active learning in rodent-borne zoonoses* takes the South Korea case study forward, from October 2026 to December 2027. I lead it from Uppsala University, with Chungnam National University as the Korean partner. Landscape embeddings from Earth-observation imagery describe habitat, Bayesian spatial models downscale host hazard to human incidence, and an active-learning step directs rodent sampling to sites where a new sample would most reduce uncertainty and that a field team can reach.

```{r}
#| results: asis
render_papers("10.1371/journal.pesy.0000023", this_page)
```

## Outbreak surveillance at UKHSA

From 2021 to 2022 I was a Senior Analyst in the Outbreak Surveillance Team at the UK Health Security Agency, working on the emergence of the SARS-CoV-2 Omicron variant and the 2022 mpox outbreak. I compiled and managed the national Omicron line list, which made the early estimates of vaccine effectiveness against Omicron possible, and contributed to the investigation of the first UK mpox cases without travel history.

```{r}
#| results: asis
render_papers(c("10.1056/nejmoa2119451", "10.1016/s1473-3099(22)00309-7", "10.1017/s0950268823000420",
                "10.2807/1560-7917.es.2022.27.11.2200143", "10.2807/1560-7917.es.2022.27.22.2200422"), this_page)
```

## mpox in endemic regions

The 2022 outbreak drew attention to a disease long under-reported where it is endemic. A commentary in the *International Journal of Infectious Diseases* proposes two drivers of the rise in outbreaks: growing synanthropic rodent populations in West and Central Africa, and a population without immunity 45 years after smallpox vaccination ended.

The reservoir of mpox virus is still unidentified. Virus has been isolated from wild animals twice, and antibodies have been found in several further species.

```{r}
#| label: mpox-table
mpox_reservoirs <- read_csv(here("monkeypox", "data", "monkeypox_reservoirs.csv"), show_col_types = FALSE)

mpox_reservoirs |>
  mutate(References = cell_spec(References, "html", link = Link)) |>
  select(-Link, -Identification) |>
  kable("html", escape = FALSE, caption = "Wild animals with evidence of mpox virus infection. Adapted from Haider et al. 2022.") |>
  pack_rows(index = table(fct_inorder(mpox_reservoirs$Identification))) |>
  kable_styling(bootstrap_options = c("hover", "condensed"), font_size = 10)
```

```{r}
#| results: asis
render_papers("10.1016/j.ijid.2022.05.058", this_page)
```

## COVID-19 modelling with CMMID

In 2020 and 2021 I was part of the COVID-19 working group of the Centre for Mathematical Modelling of Infectious Diseases at the London School of Hygiene and Tropical Medicine. The group's models informed the response in the UK and internationally: early spread and under-ascertainment, contact tracing, testing and quarantine, travel, social contacts under restrictions, and the value of vaccination.

```{r}
#| results: asis
cmmid_papers <- c(
  "10.1016/s1473-3099(20)30457-6", "10.1038/s41467-021-25531-5", "10.1186/s12916-020-01872-8", "10.7554/elife.58699",
  "10.1038/s41467-020-18783-0", "10.1016/s2468-2667(20)30308-x", "10.2807/1560-7917.es.2021.26.39.2001440",
  "10.1038/s41591-020-1036-8", "10.1038/s41467-021-22213-0", "10.1016/s1473-3099(21)00079-7", "10.1186/s12916-020-01779-4",
  "10.1186/s12916-021-01982-x", "10.1186/s12916-021-01906-9", "10.1186/s12916-021-01924-7", "10.1371/journal.pcbi.1009098",
  "10.1186/s12916-020-01790-9", "10.1038/s41467-022-32404-y", "10.2807/1560-7917.es.2021.26.49.2001809",
  "10.1017/s0950268820000424", "10.1186/s12889-022-13069-0", "10.1186/s12913-021-06509-x", "10.1038/s41598-021-86266-3")
cat(sprintf('<details class="pub-group">\n<summary>Show the %d papers</summary>\n\n', length(cmmid_papers)))
render_papers(cmmid_papers, this_page)
cat("</details>\n\n")
```

## Clinical work

From 2019 to 2022 I was a resident doctor at the Hospital for Tropical Diseases, University College London Hospitals, and worked clinically through the pandemic. With colleagues there I evaluated the 4C Mortality Score, a risk score for patients admitted with COVID-19, in UK hospitals.

```{r}
#| results: asis
render_papers("10.1101/2021.12.18.21268003", this_page)
```
