---
title: "Lassa epidemiology"
subtitle: "Case burden, reporting and the limits of the sampling record"
description: "How many Lassa fever cases there really are, and what the surveillance and sequence record can support."
categories: [Lassa Fever, Epidemiology]
toc: true
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-note appearance="minimal"}
Placeholder page. The framing text is a draft. The explainer notes for the
under-reporting and sequencing-bias papers are written and will be pulled in
from `data/paper_notes/` when the programme pages are wired to the
publications data.
:::

Reported Lassa fever case counts are the input to almost every risk model, and
they are incomplete in ways that vary by country and by year. This programme
asks what can be inferred anyway.

## Detections in hosts

Records of *Mammarenavirus lassaense* detected in small mammals across West
Africa, drawn from the Project ArHa database. The assay vocabulary is being
reconciled, so some findings currently appear twice.

```{r}
#| label: lassa-reservoirs
#| echo: false
#| message: false
#| warning: false

# --- 1. Load Packages ---
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, kableExtra, here, ggspatial, cowplot, readr)

# --- 2. Load and Prepare Data ---
# Define the path to the database directory
db_directory <- here::here("..", "arenavirus_hantavirus", "data", "database")

# Find the most recent database file in that directory
latest_db_file <- list.files(
    path = db_directory,
    pattern = "^Project_ArHa_database_.*\\.rds$",
    full.names = TRUE
  ) |>
  sort(decreasing = TRUE) |> # Sort descending, newest file will be first
  head(1) # Take the first file from the list

if (length(latest_db_file) == 0) {
  stop("No database file found in the directory: ", db_directory)
}

arha_db <- read_rds(latest_db_file)

endemic_countries <- c("Nigeria", "Sierra Leone", "Liberia", "Guinea",
                       "Benin", "Togo", "Mali", "Ghana", "Côte d'Ivoire",
                       "Senegal", "Burkina Faso")

# --- 3. Process Data to Find Earliest Evidence ---
arha_records <- arha_db$pathogen |>
  filter(pathogen_species_ncbi == "Mammarenavirus lassaense") |>
  filter(study_id != "grant_75") |>
  select(pathogen_record_id, host_record_id, pathogen_species_ncbi, number_tested, number_positive, assay) |>
  left_join(arha_db$host |>
              select(host_record_id, study_id, host_species, host_genus, gadm_country, start_date), by = "host_record_id") |>
  mutate(host_species = case_when(is.na(host_species) ~ paste0(host_genus, " spp."),
                                  TRUE ~ host_species)) |>
  filter(
    number_positive >= 1,
    gadm_country %in% endemic_countries
  ) |>
  left_join(arha_db$citations, by = "study_id") |>
  group_by(host_species, assay) |>
  slice_min(order_by = start_date, n = 1, with_ties = FALSE) |>
  ungroup()

manual_records <- tribble(
  ~study_id, ~host_species, ~host_genus, ~gadm_country, ~assay, ~publication_year, ~author, ~doi,
  "ds_402", "Mastomys natalensis", "Mastomys", "Nigeria", "Culture", 1975, "Wulff, H., et al.", "https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2366366/",
  "ds_402", "Rattus rattus", "Rattus", "Nigeria", "Culture", 1975, "Wulff, H., et al.", "https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2366366/",
  "ds_402", "Mus minutoides", "Mus", "Nigeria", "Culture", 1975, "Wulff, H., et al.", "https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2366366/",
  "ds_403", "Mastomys natalensis", "Mastomys", "Sierra Leone", "Culture", 1974, "Monath, T. P., et al.", "https://www.jstor.org/stable/1738849"
)

# --- 4. Prepare Data for Table Display ---
# This selects and formats the columns for the final table.
combined_data_for_table <- bind_rows(arha_records, manual_records) |>
  group_by(host_genus) |>
  mutate(min_year_in_genus = min(publication_year, na.rm = TRUE)) |>
  ungroup() |>
  mutate(
    host_genus = fct_reorder(host_genus, min_year_in_genus),
    author_short = str_extract(author, "[^,]+"),
    reference_text = paste(author_short, publication_year, sep = ", "),
    link = case_when(
      full_text_id == "ft_783" ~ "https://biblio.naturalsciences.be/associated_publications/bjz/bibliographic-references/139-2/bjz-139_2-p93-102.pdf",
      str_detect(doi, fixed("ezaccess.libraries.psu.edu")) ~ str_replace(doi, fixed("pubmed-ncbi-nlm-nih-gov.ezaccess.libraries.psu.edu"), "pubmed.ncbi.nlm.nih.gov"),
      !is.na(doi) & str_starts(doi, "10\\.") ~ paste0("https://doi.org/", doi),
      !is.na(doi) ~ doi, # This handles pre-existing full URLs
      TRUE ~ NA_character_
    )
  ) |>
  arrange(host_genus, host_species, publication_year) |>
  select(
    Year = publication_year,
    Country = gadm_country,
    Species = host_species,
    Assay = assay,
    References = reference_text,
    link,
    Genus = host_genus
  )

# --- 5. Generate the kableExtra HTML Table ---
# Dynamically create the index for grouping rows by genus.
tbl_groups <- table(combined_data_for_table$Genus)
names(tbl_groups) <- levels(combined_data_for_table$Genus)

combined_data_for_table |>
  mutate(References = cell_spec(References, "html", link = link)) |>
  select(-link, -Genus) |>
  kbl("html", escape = FALSE, caption = "Comprehensive list of Mammarenavirus lassaense findings in hosts across West Africa.") |>
  pack_rows(index = tbl_groups) |>
  kable_styling(bootstrap_options = c("hover", "condensed"),
                font_size = 10)

```

## Reported human cases

```{r lassa-map, echo = FALSE, message = FALSE, warning = FALSE}
lassa_raw <- read_csv(here("lassa", "data", "confirmed_cases.csv"))

lassa_clean <- lassa_raw %>%
  filter(case_when(country == "nigeria" ~ region == "all",
                   TRUE ~ confirmed_cases > 0)) %>%
  group_by(year, country) %>%
  summarise(confirmed_cases = sum(confirmed_cases)) %>%
  mutate(country = str_to_title(country),
         country = recode(country, 
                          "Cote D'ivoire" = "Côte d'Ivoire"),
         # confirmed_cases = log(confirmed_cases),
         endemic = factor(case_when(country %in% c("Benin", "Ghana", "Guinea", "Liberia", "Mali",
                                                   "Nigeria", "Sierra Leone", "Togo") ~ "",
                             TRUE ~ "Non-endemic")),
         country_label = factor(case_when(country %in% c("Benin", "Ghana", "Guinea", "Liberia", "Mali",
                                                         "Nigeria", "Sierra Leone", "Togo") ~ country,
                                          TRUE ~ "Combined"),
                                levels = c("Benin", "Ghana", "Guinea", "Liberia", "Mali",
                                           "Nigeria", "Sierra Leone", "Togo", "Combined"))) %>%
  filter(year >= 2011)

colours_wa <- c("Endemic countries" = "white",
                "Benin" = "#871a10",
                "Ghana" = "#000000",
                "Guinea" = "#be2a2c",
                "Liberia" = "#0a2864",
                "Mali" = "#50ab4a",
                "Nigeria" = "#398053",
                "Sierra Leone" = "#2f71c1",
                "Togo" = "#f7d147",
                " " = "white",
                "Non-endemic countries" = "white",
                "Côte d'Ivoire" = "#e6853b",
                "Burkina Faso" = "#107d87")

lassa_clean$country = factor(lassa_clean$country, levels=c("Endemic countries", "Benin", "Ghana", "Guinea",
                                                           "Liberia", "Mali", "Nigeria", "Sierra Leone", "Togo", " ",
                                                           "Non-endemic countries", "Côte d'Ivoire", "Burkina Faso"))

case_plot <- ggplot(lassa_clean, aes(x = year, y = confirmed_cases, fill = country)) +
  geom_col() +
  scale_x_continuous(breaks = seq(from = min(lassa_clean$year), to = max(lassa_clean$year), by = 2)) +
  scale_y_continuous(breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1))))) +
  scale_fill_manual(values = colours_wa,
                    drop=FALSE) +
  guides(fill=guide_legend(ncol=1)) +
  theme(legend.key = element_rect(fill=NA),
        legend.title = element_blank()) +
  facet_wrap(~ endemic + country_label , scales = "free_y") +
  annotate(geom = "rect", xmin = 2013, xmax = 2016, ymin = -Inf, ymax = Inf, alpha = 0.1) +
  annotate(geom = "rect", xmin = 2020, xmax = 2024, ymin = -Inf, ymax = Inf, alpha = 0.1) +
  annotate(geom = "rect", xmin = year(Sys.Date()), xmax = year(Sys.Date()) + 1, ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "orange") +
  theme_minimal() +
  labs(x = element_blank(),
       y = "Confirmed cases",
       fill = element_blank())

mastomys_data <- read_rds(here("lassa", "data", "mastomys_data.rds"))

mastomys_distribution <- ggplot() +
  geom_sf(data = mastomys_data$wa_countries, aes(fill = endemic), lwd = 0.1) +
  geom_sf_text(data = mastomys_data$wa_countries, aes(label = NAME_0), size = 5) +
  scale_fill_manual(values = c("Endemic" = "#f2aa86",
                               "Outbreaks" = "#fef8e1",
                               "No cases" = "#fafafa")) +
  geom_sf(data = mastomys_data$iucn, fill = "#C12D20", alpha = 0.4, lwd = 0.1) +
  theme_minimal() +
  labs(fill = element_blank(),
       x = element_blank(),
       y = element_blank()) +
  coord_sf(xlim = c(mastomys_data$afr_bbox[1], mastomys_data$afr_bbox[3]), ylim = c(mastomys_data$afr_bbox[2], mastomys_data$afr_bbox[4])) +
  annotation_scale(height = unit(0.1, "cm")) +
  annotation_north_arrow(style = north_arrow_minimal(text_size = 7),
                         location = "tl")

# This map has been produced using IUCN range data and country shapefiles. The code is available here: https://github.com/DidDrog11/miscellaneous/blob/main/scripts/lassa_cases.R

combined_plot <- plot_grid(case_plot, mastomys_distribution, nrow = 2,
                           labels = c("A", "B"), greedy = TRUE)
save_plot(here("lassa", "images", "lassa_cases_endemic.svg"), combined_plot, base_height = c(8, 6), base_width = c(12, 12))

```

![A) Reported human cases of Lassa fever, 2011 onwards. The first grey band (2013-2016) is the West African Ebola epidemic, the second (2020-) the COVID-19 pandemic, and the orange band an incomplete year of reporting. B) Predicted distribution of *Mastomys natalensis* over known endemic countries.](/lassa/images/lassa_cases_endemic.svg)

::: {.callout-warning appearance="minimal"}
The case figure is current to 2023-07-13 and needs regenerating.
:::
